---
title: "AmazonReviewModeling"
author: "Sofia Caceres"
date: "2026-04-21"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidytext)
library(knitr)

results_dir <- "/Users/sofiacaceres/Desktop/stats/stat405/Group12_Project/results"

rating_counts <- read.csv(file.path(results_dir, "rating_counts.csv"))
word_counts_full <- read.csv(file.path(results_dir, "word_counts_by_group.csv"))

total_reviews <- sum(rating_counts$count)

ratings_summary <- rating_counts %>%
  mutate(percent = round(100 * count / total_reviews, 2))

review_word_counts <- word_counts_full %>%
  filter(group %in% c("low", "high")) %>%
  mutate(review_label = if_else(group == "low", "bad", "good")) %>%
  select(review_label, word, n = count)

word_counts <- review_word_counts %>%
  pivot_wider(
    names_from = review_label,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(
    total = bad + good,
    bad_prop = bad / total,
    good_prop = good / total,
    difference = good_prop - bad_prop
  )
```

# Full Dataset Summary

This report uses the full Amazon review dataset processed with `Group12_Project/analyze_reviews.sh`. The original full data file has one truncated final line, so the analysis uses the complete review records.

```{r}
total_reviews
```

```{r}
ratings_summary %>%
  rename(
    Rating = rating,
    Count = count,
    Percent = percent
  ) %>%
  kable()
```

```{r}
ggplot(ratings_summary, aes(x = factor(rating), y = count, fill = factor(rating))) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  labs(
    title = "Distribution of Review Ratings in the Full Amazon Dataset",
    x = "Rating",
    y = "Count"
  )
```

# Word Counts

For the good/bad comparison, 1-star and 2-star reviews are labeled bad, while 4-star and 5-star reviews are labeled good. Three-star reviews are counted in the rating distribution but excluded from the good/bad word comparison.

```{r}
top_words <- review_word_counts %>%
  group_by(review_label) %>%
  slice_max(n, n = 15) %>%
  ungroup()

ggplot(top_words, aes(x = reorder_within(word, n, review_label), y = n, fill = review_label)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ review_label, scales = "free") +
  scale_x_reordered() +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  coord_flip() +
  labs(
    title = "Most Common Words in Good vs Bad Amazon Reviews",
    x = "Word",
    y = "Count"
  )
```

```{r}
word_counts %>%
  filter(total >= 1000) %>%
  arrange(difference) %>%
  head(20)

word_counts %>%
  filter(total >= 1000) %>%
  arrange(desc(difference)) %>%
  head(20)
```

```{r}
bad_words <- word_counts %>%
  filter(total >= 1000) %>%
  arrange(difference) %>%
  head(20)

good_words <- word_counts %>%
  filter(total >= 1000) %>%
  arrange(desc(difference)) %>%
  head(20)

bad_words
good_words
```

```{r}
plot_words <- bind_rows(
  bad_words %>% mutate(group = "More common in bad reviews"),
  good_words %>% mutate(group = "More common in good reviews")
)

ggplot(plot_words, aes(x = reorder(word, difference), y = difference, fill = group)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ group, scales = "free_y") +
  coord_flip() +
  labs(
    title = "Words Most Associated with Good vs Bad Reviews",
    x = "Word",
    y = "Good proportion minus bad proportion"
  )
```

# Modeling Note

The earlier logistic regression and random forest sections created document-level matrices from the 10,000-review sample. The full dataset has more than six million complete records, so this version uses the full-corpus rating counts and word-count summaries generated by streaming the data instead of loading the entire JSON file into memory.
