---
title: "Consumer Behavior & Product Placement Analysis"
subtitle: "by Sofia Caceres"
output: html_document
date: "2023-11-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,
                      warning = FALSE, error = TRUE, fig.height = 3)
library(tidyverse)
library(lubridate)
source("../scripts/viridis.R")
source("../scripts/ggprob.R")
theme_set(theme_minimal())
```


```{r}
library(readr)
cereal = read.csv("/Users/sofiacaceres/Desktop/stats/stats240/project/cereal.csv")
```

**Introduction**


The U.S. cereal industry uses more than 400 million kilograms of sugar annually[3]. Product ratings and shelf placement are important indicators of consumer demand and retail strategy, making them useful measures for understanding how product characteristics influence purchasing behavior. This project examines whether sugar content is associated with cereal popularity and shelf placement using nutritional and product-rating data from major cereal brands. The analysis evaluates whether cereals with higher sugar content are more likely to occupy prominent shelf positions and receive stronger consumer ratings. Results indicate a statistically significant relationship between sugar content and shelf placement, suggesting that product positioning is not random with respect to sugar levels.


**Background**

Cereals, an indispensable breakfast choice in daily life, derive their name from Ceres, the ancient Roman goddess of agriculture. In the United States, about 50% of people start their day with cereal every morning. Cereals are not only a popular breakfast option but also a rich source of fiber, which plays an important role in maintaining good health.

The primary objective of this analysis is to provide insight into the nutritional composition of commonly consumed breakfast cereals. This enables a variety of analyses, including evaluating nutritional value and ingredient composition, which may influence dietary choices and health considerations.

The impact of shelf placement on cereal products is significant, particularly in terms of consumer visibility and purchasing behavior. Products placed at eye level (levels 2 and 3) are more likely to be noticed by adults, while lower shelves (level 1) often attract children's attention due to natural visual habits and attention patterns. Marketing strategies frequently place sweeter cereals targeted toward children on lower shelves and healthier options at adult eye level. As a result, consumer choices can be influenced by shelf placement, as more visible products tend to generate greater interest.

However, most surveyed consumers supported shelf placement as a merchandising strategy (265 respondents, 88%) while noting that brand preference and price were more influential determinants of purchasing decisions.


  The "80 Cereals" dataset contains comprehensive information on 80 different cereal products from seven manufacturers, including **American Home Food Products**, **General Mills**, **Kellogg's**, **Nabisco**, **Post**, **Quaker Oats**, and **Ralston Purina**. This dataset includes various nutritional and other relevant data points for each cereal, such as:

- **Name**: Name of the cereal.
- **Manufacturer (mfr)**: The company that produces the cereal, coded with letters representing different manufacturers.
- **Type**: Whether the cereal is cold or hot.
- **Calories**: Calories per serving.
- **Protein**: Grams of protein per serving.
- **Fat**: Grams of fat per serving.
- **Sodium**: Milligrams of sodium per serving.
- **Fiber**: Grams of dietary fiber per serving.
- **Carbohydrates (carbo)**: Grams of complex carbohydrates per serving.
- **Sugars**: Grams of sugars per serving.
- **Potassium (potass)**: Milligrams of potassium per serving.
- **Vitamins**: Vitamins and minerals content, indicated as 0, 25, or 100, representing the typical percentage of FDA recommended.
- **Shelf**: Display shelf number (1, 2, or 3, counting from the floor).
- **Weight**: Weight in ounces of one serving.
- **Cups**: Number of cups in one serving.
- **Rating**: A rating of the cereals, possibly from Consumer Reports.

  The dataset provides valuable insights into the nutritional content of popular breakfast cereals, enabling analyses regarding health, dietary choices, and consumer preferences. The data was collected by the Section of Statistical Graphics, when they sponsored a special exposition to make the data set availible to the public. All of the data collected is based on the information availible on the F&DA food label, as well as nutrition information released by the cereal companies.

  A single row of the data set represents all of the data available for that specific type and brand of cereal in the data set. It represents the cereal’s profile, both considering nutrients and popularity.

Manufacturers are represented by their first initial: <br>
A = American Home Food Products <br>
G = General Mills <br>
K = Kelloggs <br> 
N = Nabisco <br> 
P = Post <br>
Q = Quaker Oats <br>
R = Ralston Purina

  For the rest of the report, the intention would be to analyze this data to draw conclusions about cereal nutrition and its implications on health and consumer choices. This include exploring correlations between different nutritional factors and the overall rating, investigating trends among different manufacturers, and identifying any unusual factors that could affect interpretation, such as outliers or anomalies in the data.


**This graph illustrates the distribution and variability of cereal ratings across different cereal products.**

```{r}

ggplot(cereal, aes(x = rating)) + geom_boxplot(col = "pink", fill = "lightblue") + 
  labs(title = "Average Ratings of Cereals",
       x = "Rating")

```

**This graph illustrates the distribution of sugar content across cereal rating categories**

```{r}

ggplot(cereal, aes(x = sugars, group = sugars, fill = sugars)) + geom_bar() + 
  labs(title = "Average Ratings of Cereals by Sugar",
       x = "Rating",
       y = "Sugars")

```

**This visualization examines the relationship between cereal shelf placement and consumer ratings, highlighting how ratings vary across different shelf levels.**

```{r}
summary = cereal %>% 
  group_by(mfr) %>%
  summarize(n = n(),
            avg_sugar = mean(sum(sugars))/n, 
            avg_rating = mean(sum(rating))/n,
            avg_shelf_height = mean(sum(shelf))/n,
            shelf = shelf) %>%
  ungroup(mfr) %>%
  group_by(shelf) %>%
  arrange(desc(avg_sugar))

ggplot(cereal, 
       aes(x = shelf, y = rating, 
           group = shelf, fill = shelf)) + 
  geom_boxplot() + 
  labs(x = "Shelf Height", y = "Rating", title = "Shelf Height VS Rating of Cereal")
```

**This graph compares cereal sugar content with consumer ratings to examine the relationship between sweetness and product popularity.**

```{r}

ggplot(cereal, aes(x=rating, y=sugars)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Rating", y = "Sugars", title = "Amount of Sugar VS Cereal Rating")

```

**This graph examines the relationship between carbohydrate content and cereal ratings, illustrating how consumer ratings vary across different carbohydrate levels.**

```{r}
ggplot(cereal, aes(x=rating, y = carbo)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Rating", y = "Carbohydrates", title = "Number of Carbs VS Cereal Rating")

```

**This graph compares the amount of protien in a cereal box to the cereal's overall rating.**

```{r}
ggplot(cereal, aes(x=protein, y = rating, fill = protein)) +
  geom_col() +
  labs(x = "Protein", y = "Rating", title = "Amount of Protein VS Cereal Rating")

```

**These visualizations provide a manufacturer-level view of the data, highlighting differences in performance across cereal brands.**

**Comparing the average cereal rating across manufacturers, identifying brands associated with higher consumer ratings**

```{r}

cereal$mfr = factor(cereal$mfr,
                        levels = c('A', 'G', 'K', 'N', 'P', 'Q', 'R'),
                        labels = c('American Home Food Products', 'General Mills',
                                   'Kelloggs', 'Nabisco', 'Post', 'Quaker Oats',
                                   'Ralston Purina'))

mean_ratings = cereal %>%
  group_by(mfr) %>%
  summarise(mean_rating = mean(rating),
            mean_sugar = mean(sugars))

ggplot(mean_ratings, aes(x=mfr, y=mean_rating, fill=mfr)) +
  geom_bar(stat='identity') +
  theme_minimal() +
  labs(title='Mean Consumer Ratings of Cereal by Manufacturer',
       x='Manufacturer',
       y='Mean Rating') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```

**Similarly, this graph compares the average sugar content across cereal manufacturers, highlighting differences in product formulation among brands.**

```{r} 
ggplot(mean_ratings, aes(x=mfr, y=mean_sugar, fill=mfr)) +
  geom_bar(stat='identity') +
  theme_minimal() +
  labs(title='Mean Sugars of Cereal by Manufacturer',
       x='Manufacturer',
       y='Mean Rating')  

```

**Analysis**

```{r}
cereal_sum = cereal %>%
  group_by(shelf) %>%
  mutate(mfr = case_when(mfr == "A" ~ "American Home Food Products", 
                         mfr == "G" ~ "General Mills",
                         mfr == "K" ~ "Kelloggs",
                         mfr == "N" ~ "Nabisco",
                         mfr == "P" ~ "Post",
                         mfr == "Q" ~ "Quaker Oats",
                         mfr == "R" ~ "Ralston Purina")) %>%
  summarize(n = n(),
            mean = mean(sugars),
            sd = sd(sugars))

cereal_plot = cereal %>% group_by(shelf) %>%
  summarise(shelf = shelf,
            sugars = sugars)

ggplot(cereal_plot, aes(x = shelf, y = sugars, fill = shelf, group = shelf)) + 
  geom_boxplot() + 
  geom_point(position = position_jitter(width=0.4, height=0)) + 
  theme_minimal() +
  labs(x = "Shelf Height", y = "Sugar in Grams", title = "Cereal Sugar Content by Shelf Height")

x = cereal_plot %>% 
  filter(shelf == 1 | shelf == 3) %>%
  pull(sugars)

y = cereal_plot %>%
  filter(shelf == 2) %>%
  pull(sugars)

t_test = t.test(y,x)
  
dof = 36.613 #t.test(x, y)$parameter
mean_x=mean(x)
s_x = sd(x)
n_x = length(x)

mean_y = mean(y)
s_y = sd(y) 
n_y = length(y)

ci = mean_x - mean_y + c(-1,1)*qt(0.975, dof)*sqrt(s_x^2/n_x + s_y^2/n_y)
ci

## t.test(x, y, var.equal = TRUE) #assumes equal variances
## ci #does not assume equal variances

## The results of the t-test indicate that the average sugar content of cereals on shelf 2 is significantly higher than the average sugar content of cereals on shelves 1 and 3, which is in line with the graphical display of the data.

## These results suggest that the shelf height of cereals with more sugar is higher than the average sugar content of cereals across all shelf heights. This information can be used by marketers and businesses to design marketing strategies aimed at targeting customers who prefer cereals with a higher sugar content.

welch_df = function(x, y)
{
  sx = sd(x)
  sy = sd(y)
  nx = length(x)
  ny = length(y)
  
  df = (sx^2/nx + sy^2/ny)^2 / (sx^4/(nx^2*(nx-1)) + sy^4/(ny^2*(ny-1)))
  
  return ( df )
}
```

- State Hypotheses
$H_0: \mu_1 = \mu_2$    
$H_a: \mu_1 \neq \mu_2$

- Test statistic
$t = \frac{\bar{x} - \bar{y}}{\text{SE}(\bar{x} - \bar{y})}$

```{r}

t.test(x,y)

## manual
dof = 36.613 #t.test(x, y)$parameter
n_x = length(x)
n_y = length(y)
se = sqrt(var(x)/n_x + var(y)/n_y)
tstat = (mean(x) - mean(y)) / se
tstat
pvalue = 2*pt(-abs(tstat), dof) 
pvalue
t.test(x,y)$p.value

```

To evaluate the relationship between shelf placement and sugar content, an inference for two means approach was used. This method allows for comparison of average sugar content across shelf levels while accounting for differences in the number of cereal products placed on each shelf.

Results indicate with 95% confidence that cereals placed on the second shelf contain, on average, between 5.64 and 9.62 more grams of sugar than cereals placed on shelves one and three.

The analysis also provides strong statistical evidence that cereals on the second shelf have a higher mean sugar content than cereals on the other shelf levels. This conclusion is supported by a two-sided t-test with a p-value of 0.00063, which is well below the 0.05 significance threshold.


**Discussion**

In the very beginning of the exploration, various graphs were created to assess different nutrients and how they relate to cereal rankings. After exploring the data with nutrients such as carbohydrates and protein, a significant relationship was identified between sugar content and cereal ranking.

In order to draw conclusions about a cereal's popularity, the average consumer ranking was compared to the shelf on which the cereal is found in the grocery store. Shelf 2 was found to have the highest ranking among consumers. The total amount of sugar on each shelf was then compared to determine which shelf contained the most sugar. Combining these findings showed that Shelf 3 had the highest amount of sugar, while Shelf 2 had the highest consumer ranking. This result does not support the original hypothesis, as the highest amount of sugar did not correspond to the most popular cereals. Findings also showed that cereals with lower amounts of sugar received higher rankings among consumers and were more commonly found on Shelf 2.

There are two shortcomings in this dataset. The first is that the dataset has not been updated in 6 years. As time progresses, Americans have increased their daily calorie intake and their daily calories from sugar. In the past 6 years, as sugary calorie intake has increased, consumer rankings of sugary cereals may have increased as well [5]. Another shortcoming is the exclusion of certain cereal brands. For example, Reese's Puffs cereal is not included in the dataset. The absence of popular cereal brands may affect the accuracy and generalizability of the findings.

New questions for this dataset could include focusing on different nutrients. Future analyses could examine how the amount of vitamins or sodium affects a cereal's popularity among consumers.

A new approach to answering this question could involve finding a more current and comprehensive dataset or modifying the statistical methods used to draw conclusions. Similar techniques could be applied, including numerical summaries, confidence intervals, statistical modeling, and interpretation of results when populations are of equal size. A revised dataset could also be created with unevenly distributed groups and analyzed through manual confidence interval calculations. These findings could then be evaluated through additional hypothesis testing to strengthen the conclusions.

In the past six years, many new cereal brands have entered the market. To refine the understanding of cereal popularity in relation to sugar content, future analyses could include more recently introduced cereals and brands that were not available when this dataset was created. As consumer preferences evolve, incorporating newer products may provide a more accurate representation of current purchasing behavior.

The exploration of cereal popularity and its relationship with sugar content initially focused on analyzing several nutrients, particularly sugar, carbohydrates, and protein. Contrary to the original hypothesis, cereals with lower sugar content tended to receive higher consumer rankings and were more commonly found on Shelf 2 in grocery stores. Despite Shelf 3 having the highest sugar content, Shelf 2 remained the most popular among consumers.



**References**


[1] - Cereal data was obtained from a comprehensive dataset available on Kaggle^[See the dataset at [Kaggle Cereal Dataset](https://www.kaggle.com/datasets/crawford/80-cereals/data)].

[2] - Additional cereal data was referenced from the CMU dataset^[Details can be found at [CMU Dataset](https://lib.stat.cmu.edu/datasets/1993.expo/)].

[3] - Historical information and interesting facts about cereals were gathered from a dedicated cereal history website^[More information available at [History of Cereals](http://www.historyofcereals.com/cereal-facts/interesting-facts-about-cereals/)].

[4] - The effect of a shelf placement intervention on sales of healthier and less healthy breakfast cereals in supermarkets: A co-designed pilot study
^[Science Direct [Shelf Placement](https://www.sciencedirect.com/science/article/pii/S0277953620305566)].

[5] - Americans are consuming far more calories each day than is recommended (daily intake should be around 2,000 calories for women and 2,500 for men)^[Business Insider [Average American Calorie Intake](https://www.businessinsider.com/daily-calories-americans-eat-increase-2016-07#daily-calories-from-sugar-and-artificial-sweeteners-has-also-risen-by-almost-100-3)].
